Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-12-04 WholePlotTrajectories.Rmd
intro

The large areas covered with tropical forests worldwide hold crucial environmental, economic and social values. 
They provide wood and multiple non-timber forest products, shelter a diversified fauna, regulate the local and regional climates, determine the cycles of carbon, water and nutrient, and ensure cultural and human well-being. 
The growing demand in forests products together with current global changes increases the pressure on remaining natural forests [@Morales-Hidalgo2015] and threatens the maintenance and dynamics in space and time of community structure, composition and functioning [@Anderson-Teixeira2013].

In tropical forests, ecological communities are regularly re-shaped by natural disturbance events changing both the abiotic environment, through the fluxes of light, heat and water [@Goulamoussene2017], and the biotic interactions such as competition among species [@Chesson2000]. 
One of the cornerstone of tropical forest ecology is to understand the processes and drivers of ecosystems response to such disturbance [@Chazdon2003a].
This response has been largely studied through forest structural parameters such as aboveground biomass, tree height or stem density [@Piponiot2016;@Rutishauser2016]. 
Thereafter, models based on the observed trajectories after disturbance gave important insights into the recovery of ecosystem processes and services [@Herault2018]. 
Regarding community diversity and composition, however, post-disturbance trajectories are not as thoroughly understood.
Several studies and meta-analyses highlighted determined successions after disturbance, mainly driven by the recruitment of pioneer and heliophilous species [@Carreno-Rocabado2012;@Poorter2016].
Community trajectories thus proved to rely on deterministic processes and post-disturbance responses would be predictable according to the disturbance intensity and the comunity diversity and composition.
Still, observed trajectories in diversity and composition often deviate from predictions and the underlying ecological processes are debated [@Norden2015].
Deeper understanding of tropical forest response to disturbance could be provided by long term monitoring, to complete the conclusions often based on single decades monitoring [@Carreno-Rocabado2012], dynamic trajectories to complete chronosequence studies [@Chazdon2007], and through a complete set of taxonomic and functional composition and diversity metrics to complete studies based only on taxonomic richness or functional composition [@Carreno-Rocabado2012,@Mouillot2013].

Deterministic processes are thus recognized as important drivers of community trajectories.
This is formalized by the intermediate disturbance hypothesis (IDH), predicting a miximum species diversity after moderate disturbance intensity [@Molino2001;@Kariuki2006a].
Validations of the IDH though remain scarce in the long-term and mainly rely on the analysis of taxonomic richness [@Molino2001]. Taxonomic richness alone, however, gives limited or misleading information on forests recovery and functioning [@Chaudhary2016]. More ecological-meaningful analysis would couple richness with (i) evenness, which would reveal the changes in the species abundance distribution and thus the underlying ecological processes, and (ii) composition, which is crucial for conservation issues [@Lavorel2002;@Bellwood2006]. Furthermore, a functional approach accounting for species biological attributes would directly link community diversity, composition and redundancy to ecosystem functioning and to its environmental constraints [@Violle2007b;@Baraloto2012a].
In that respect, the functional trait-based approach that focus on major traits related to species ecology and mediate species performance in a given environment was successfully adopted [@Diaz2005].
For instance, the functional approach revealed in tropical rainforests the deterministic processes entailing, after disturbance, a functional shift from a dominance of "conservative" slow-growing species dealing with scarce resources to "acquisitive" fast-growing species with rapid and efficient use of abundant resources [@Reich2014;@Herault2011].
This shift is translated into the trajectories of key functional traits related to resource acquisition (leaf and stem traits) and life-history traits (seed mass, maximum size) [@Wright2004;@TerSteege2006;@Westoby2006a;@Chave2009b].
Eventually, a complete overview of community response to disturbance would encompass the changes in functional redundancy, that quantifies the amount of shared trait values among species [@Carmona2016].
The high functional redundancy of hyper-diverse tropical forests [@Bellwood2006] mitigates the impacts of species removal on ecosystem functioning and determines the resilience of communities after disturbance [@Elmqvist2003; @Diaz2005].

>>>>

The large areas covered with tropical forests worldwide hold crucial environmental, economic and social values. 
They provide wood and multiple non-timber forest products, shelter a diversified fauna, regulate the local and regional climates, determine the cycles of carbon, water and nutrient, and ensure cultural and human well-being. 
The growing demand in forests products together with current global changes increases the pressure on remaining natural forests [@Morales-Hidalgo2015].
In this context, the maintenance of tropical forest community structure, composition and functioning is increasingly threatened [@Anderson-Teixeira2013].

Tropical forests are subject to natural disturbance events, specifically tree fall gaps, determining and maintaining the characteristics of ecosystems [@Schnitzer2001]. 
Disturbances change both the biotic and abiotic environment, through modifications of the fluxes of light, heat and water [@Goulamoussene2017], and of the interactions among species such as competition [@Chesson2000].
Understand the drivers of community response to disturbance and the underlying ecological processes is then crucial to anticipate the fate of tropical forests in the changing context [@Chazdon2003a].
Community response has been largely studied through forest structural parameters such as aboveground biomass, tree height or stem density [@Piponiot2016;@Rutishauser2016]. 
Thereafter, models based on the observed trajectories after disturbance gave important insights into the recovery of ecosystem processes and services [@Herault2018]. 
Regarding community diversity and composition, however, post-disturbance trajectories are not as thoroughly understood.
It is recognized that determined succession, mainly driven by the recruitment of pioneer and heliophilous species, shape post-disturbance trajectories [@Carreno-Rocabado2012;@Poorter2016].
Community trajectories would then rely on deterministic processes and would be predictable according to the disturbance intensity and to the characteristics of the pre-disturbance community.
Yet, observed trajectories in diversity and composition often deviate from predictions.
Besides, although the deterministic ecological processes at stake are identified, their relative importance compared to stochasticity remains debated [@Norden2015].
This debate translates into the validation and the extent of the intermediate disturbance hypothesis (IDH).
The IDH assumes the predominance of deterministic processes, and predicts a maximum species diversity after moderate disturbance intensity [@Molino2001;@Kariuki2006a].
Validations of the IDH remain scarce in the long-term and often rely on the analysis of taxonomic richness [@Molino2001]. 
Taxonomic richness alone, however, gives limited or misleading information on forests recovery and functioning [@Chaudhary2016]. 
More ecological-meaningful analysis would couple richness with species evenness, which reveals the changes in the species abundance distribution, and (ii)species composition, which is crucial for conservation issues [@Lavorel2002;@Bellwood2006].
Furthermore, a functional approach accounting for species biological attributes would directly link community diversity, composition and redundancy to ecosystem functioning and to its environmental constraints [@Violle2007b;@Baraloto2012a].
In that respect, the functional trait-based approach that focus on major traits related to species ecology and mediate species performance in a given environment was successfully adopted [@Diaz2005].
For instance, the functional approach revealed in tropical rainforests the deterministic processes entailing, after disturbance, a functional shift from a dominance of "conservative" slow-growing species dealing with scarce resources to "acquisitive" fast-growing species with rapid and efficient use of abundant resources [@Reich2014;@Herault2011].
This shift is translated into the trajectories of key functional traits related to resource acquisition (leaf and stem traits) and life-history traits (seed mass, maximum size) [@Wright2004;@TerSteege2006;@Westoby2006a;@Chave2009b].
Eventually, a complete overview of community response to disturbance would encompass the changes in functional redundancy, that quantifies the amount of shared trait values among species [@Carmona2016].
The high functional redundancy of hyper-diverse tropical forests [@Bellwood2006] mitigates the impacts of species removal on ecosystem functioning and determines the resilience of communities after disturbance [@Elmqvist2003; @Diaz2005].

Improve the understanding of diversity and composition response to disturbance would then require to combine taxonomic and functional diversity and composition metrics that is not often provided [@Carreno-Rocabado2012,@Mouillot2013]. 
Long term monitoring provided trajectories over several decades would allow to put in perspectives the existing studies that often correspond to chronosequence studies [@Chazdon2007] or cover 10 or 15 years after disturbance [@Carreno-Rocabado2012].

### 2018-12-04 WholePlotTrajectories.tex
intro

The large areas covered with tropical forests worldwide hold crucial
environmental, economic and social values. They provide wood and
multiple non-timber forest products, shelter a diversified fauna,
regulate the local and regional climates, the carbon, water and nutrient
cycles, and ensure cultural and human well-being. The growing demand in
forests products together with current global changes increases the
pressure on remaining natural forests \citep{Morales-Hidalgo2015} and
threatens the maintenance and dynamics in space and time of community
structure, composition and functioning \citep{Anderson-Teixeira2013}.

In tropical forests, ecological communities are regularly re-shaped by
natural disturbance events changing both the abiotic environment,
through the fluxes of light, heat and water \citep{Goulamoussene2017},
and the biotic interactions such as competition among species
\citep{Chesson2000}. One of the cornerstone of tropical forest ecology
is to understand the processes and drivers of ecosystems response to
disturbance \citep{Chazdon2003a}. For now, this has been largely studied
through forest structural parameters such as aboveground biomass, tree
height or stem density \citep{Piponiot2016, Rutishauser2016} that are
rapid and convenient to measure. These structural parameters have been
successfully modeled, giving important insights into the recovery of
ecosystem processes and services \citep{Herault2018}. However the
response of forests diversity and composition remains unclear, albeit it
determines the productivity, stability and functioning of ecosystems
\citep{Tilman2014, Liang2016}. In the short-term, moderate disturbance
may lead to positive impacts on community diversity, an idea formalized
by the intermediate disturbance hypothesis (IDH) which predicts a
maximum species diversity when disturbance intensity is not too high
\citep{Molino2001, Kariuki2006a}.

Validations of the IDH though remain scarce in the long-term and mainly
rely on the analysis of taxonomic richness \citep{Molino2001}. Taxonomic
richness alone, however, gives limited or misleading information on
forests recovery and functioning \citep{Chaudhary2016}. More
ecological-meaningful analysis would couple richness with (i) evenness,
which would reveal the changes in the species abundance distribution and
thus the underlying ecological processes, and (ii) composition, which is
crucial for conservation issues \citep{Lavorel2002, Bellwood2006}.
Furthermore, a functional approach accounting for species biological
attributes would directly link community diversity, composition and
redundancy to ecosystem functioning and to its environmental constraints
\citep{Violle2007b, Baraloto2012a}. In that respect, the functional
trait-based approach that focus on major traits related to species
ecology and mediate species performance in a given environment was
successfully adopted \citep{Diaz2005}. For instance, the functional
approach revealed in tropical rainforests the deterministic processes
entailing, after disturbance, a functional shift from a dominance of
``conservative'' slow-growing species dealing with scarce resources to
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{Reich2014, Herault2011}. This shift is
translated into the trajectories of key functional traits related to
resource acquisition (leaf and stem traits) and life-history traits
(seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}. Eventually,
a complete overview of community response to disturbance would encompass
the changes in functional redundancy, that quantifies the amount of
shared trait values among species \citep{Carmona2016}. The high
functional redundancy of hyper-diverse tropical forests
\citep{Bellwood2006} mitigates the impacts of species removal on
ecosystem functioning and determines the resilience of communities after
disturbance \citep{Elmqvist2003, Diaz2005}.

@@

Because of the variability in the precision of botanical identification
efforts, we were constraint to conduct the taxonomic composition and
diversity analysis at the genus level. Taxonomic and functional
trajectories of community composition were followed in a two-dimensional
NMDS ordination space. Two NMDS using abundance-based (Bray-Curtis)
dissimilarity measures were conducted to map either taxonomic or
functional composition, the latter based on the seven leaf, stem and
life history traits (without seed mass classes). Trajectories along time
were reported through the Euclidean distance between the target
inventories and the reference inventories in 1989, \emph{i.e.} 2 years
after disturbance. Univariate trajectories of the leaf, stem and
life-history traits were also visualized with the community weighted
means (CWM) \citep{Diaz2007}. Species seed mass were given in 5 mass
classes. Seed mass trajectories were therefore reported as the
proportion of each class in the inventories (Supplementary materials).

The taxonomic diversity was reported through species richness and the
Hill number translation of the Simpson index \citep{Hill1973}. The
comparison between richness and Simpson diversity assessed the
variations in taxonomic evenness. Both indices belong to the set of HCDT
or generalized entropy, respectively corresponding to the 0 and 2 order
of diversity (q), recommended for diversity studies \citep{Marcon2015b}.
The functional diversity was reported using the functional richness and
functional evenness, \emph{i.e} the Rao index of quadratic entropy which
combines species abundance distribution and average pairwise
dissimilarity based on species functional traits.

@@

Functional redundancy was measured as the overlap among species in
community functional space \citep{Carmona2016}. First, the individuals
of the trait database were mapped in the plane of the two first axes
from a PCA analysis. Then, for each species, the traits probability
density (TPD) were computed through multivariate kernel density
estimators. Second, for each community, the functional traits
probability density were computed as the sum of the TDP of the
inventoried species weighted by species abundance. Third, the whole
functional space was divided into a 100 x 100 grid, before counting for
each cell the number of species which TPD is positive. The average count
across cells, minus 1, gives the average number of species in the
community that share the same traits, which is the Community Functional
Redundancy.

@@

\caption{Plot trajectories in terms of taxonomic composition (\textbf{(a)} and \textbf{(c)}) and functional composition (\textbf{(b)} and \textbf{(d)}) in a two-dimensional NMDS space. Lower panels (\textbf{(c)} and \textbf{(d)}) represent the Euclidean distance to initial condition along the 30 sampled years. Shaded areas are the credibility intervals.}\label{fig:NMDSplans}
\end{figure*}

@@

The second-degree polynomial regressions between (i) the percentage AGB
loss and (ii) taxonomic and functional diversity after 10, 20 and 30
years best predicted the hump-shaped curve of the disturbance impact
along the disturbance intensity gradient \ref{fig:IDHplot}. The
relationship between the disturbance impact and its intensity was more
markedly hump-shaped for the taxonomic richness than for the Simpson
diversity. For both functional richness and Rao diversity the
relationship was almost linear. The regression model better predicted
the functional richness and Rao diversity
(\(0.55<R^2_{Functional Richness}<0.72\), and
\(0.60<R^2_{Functional Rao}<0.81\)) than the taxonomic richness and
evenness (\(0.21<R^2_{Taxonomic Richness}<0.4\), and
\(-0.15<R^2_{Taxonomic Simpson}<0.43\) respectively)

@@

The taxonomic differences among local communities were already apparent
before disturbance, as revealed by the distinct starting points on the
NMDS axis 2. These differences were maintained throughout recovery
trajectories.\\
More than commonly thought, post-disturbance trajectories depended on
community initial composition, which partly determined the pool of
recruited species and constrained the trajectories towards the initial
composition. The high resilience of community taxonomy, in the sense of
the recovery of initial state, revealed that species not belonging to
the pre-disturbance community were rarely recruited because of the
commonness of dispersal limitation among tropical tree species
\citep{Svenning2005}.

@@

The IDH predicted well the enhancement of taxonomic richness after
disturbance, until an intensity threshold (20-25\% AGB loss). It also
explained the impact on the Simpson diversity and hence on community
taxonomic evenness. Taxonomic evenness was somewhat decoupled from the
disturbance intensity as already observed in the Guiana Shield
\citep{Baraloto2012a} and in Bornean tropical forests
\citep{Cannon1998}. Disturbance intensity determined the balance in the
community between pre-disturbance surviving trees and those recruited
afterward. The pool of true pioneer species specifically recruited after
disturbance is restricted in the Guiana Shield to a few common genera
(e.g. \emph{Cecropia} spp., \emph{Vismia} spp.) \citep{Guitet2018}.
Below the intensity threshold the size of the surviving community
maintained the pre-disturbance high taxonomic richness while the
recruitment of pioneers, infrequent or absent before disturbance,
increased both community taxonomic richness and evenness. Beyond the
intensity threshold, the disturbance decreased the taxonomic richness of
surviving trees which was not offset by the enrichment of pioneers, so
that the overall community taxonomic richness decreased according to the
disturbance intensity \citep{Molino2001}. An intermediate disturbance
hence increased the availability of resources and created opportunities
for pioneers, without implying too important loss of shade-tolerant
species and without preventing their maintenance in the community
\citep{Bongers2009}. For community taxonomic evenness the disturbance
impact was similar but milder, as the evenness is less sensitive to the
loss of rare species. Taxonomic evenness trajectories, below the
intensity threshold, rather represented the increasing dominance of
pioneers that balanced the usual hyper-dominance of a few species in
tropical forests. Beyond the intensity threshold, however, pioneers
became in turn highly dominant and decreased the overall evenness
\citep{Baraloto2012a}.

>>>>

The large areas covered with tropical forests worldwide hold crucial
environmental, economic and social values. They provide wood and
multiple non-timber forest products, shelter a diversified fauna,
regulate the local and regional climates, determine the cycles of
carbon, water and nutrient, and ensure cultural and human well-being.
The growing demand in forests products together with current global
changes increases the pressure on remaining natural forests
\citep{Morales-Hidalgo2015}. In this context, the maintenance of
tropical forest community structure, composition and functioning is
increasingly threatened \citep{Anderson-Teixeira2013}.

Tropical forests are subject to natural disturbance events, specifically
tree fall gaps, determining and maintaining the characteristics of
ecosystems \citep{Schnitzer2001}. Disturbances change both the biotic
and abiotic environment, through modifications of the fluxes of light,
heat and water \citep{Goulamoussene2017}, and of the interactions among
species such as competition \citep{Chesson2000}. Understand the drivers
of community response to disturbance and the underlying ecological
processes is then crucial to anticipate the fate of tropical forests in
the changing context \citep{Chazdon2003a}. Community response has been
largely studied through forest structural parameters such as aboveground
biomass, tree height or stem density
\citep{Piponiot2016, Rutishauser2016}. Thereafter, models based on the
observed trajectories after disturbance gave important insights into the
recovery of ecosystem processes and services \citep{Herault2018}.
Regarding community diversity and composition, however, post-disturbance
trajectories are not as thoroughly understood. It is recognized that
determined succession, mainly driven by the recruitment of pioneer and
heliophilous species, shape post-disturbance trajectories
\citep{Carreno-Rocabado2012, Poorter2016}. Community trajectories would
then rely on deterministic processes and would be predictable according
to the disturbance intensity and to the characteristics of the
pre-disturbance community. Yet, observed trajectories in diversity and
composition often deviate from predictions. Besides, although the
deterministic ecological processes at stake are identified, their
relative importance compared to stochasticity remains debated
\citep{Norden2015}. This debate translates into the validation and the
extent of the intermediate disturbance hypothesis (IDH). The IDH assumes
the predominance of deterministic processes, and predicts a maximum
species diversity after moderate disturbance intensity
\citep{Molino2001, Kariuki2006a}. Validations of the IDH remain scarce
in the long-term and often rely on the analysis of taxonomic richness
\citep{Molino2001}. Taxonomic richness alone, however, gives limited or
misleading information on forests recovery and functioning
\citep{Chaudhary2016}. More ecological-meaningful analysis would couple
richness with species evenness, which reveals the changes in the species
abundance distribution, and (ii)species composition, which is crucial
for conservation issues \citep{Lavorel2002, Bellwood2006}. Furthermore,
a functional approach accounting for species biological attributes would
directly link community diversity, composition and redundancy to
ecosystem functioning and to its environmental constraints
\citep{Violle2007b, Baraloto2012a}. In that respect, the functional
trait-based approach that focus on major traits related to species
ecology and mediate species performance in a given environment was
successfully adopted \citep{Diaz2005}. For instance, the functional
approach revealed in tropical rainforests the deterministic processes
entailing, after disturbance, a functional shift from a dominance of
``conservative'' slow-growing species dealing with scarce resources to
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{Reich2014, Herault2011}. This shift is
translated into the trajectories of key functional traits related to
resource acquisition (leaf and stem traits) and life-history traits
(seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}. Eventually,
a complete overview of community response to disturbance would encompass
the changes in functional redundancy, that quantifies the amount of
shared trait values among species \citep{Carmona2016}. The high
functional redundancy of hyper-diverse tropical forests
\citep{Bellwood2006} mitigates the impacts of species removal on
ecosystem functioning and determines the resilience of communities after
disturbance \citep{Elmqvist2003, Diaz2005}.

Improve the understanding of diversity and composition response to
disturbance would then require to combine taxonomic and functional
diversity and composition metrics that is not often provided
\citep[\citet{Mouillot2013}]{Carreno-Rocabado2012}. Long term monitoring
provided trajectories over several decades would allow to put in
perspectives the existing studies that often correspond to
chronosequence studies \citep{Chazdon2007} or cover 10 or 15 years after
disturbance \citep{Carreno-Rocabado2012}.

@@

Because of the variability in the precision of botanical identification
efforts, we were constraint to conduct the taxonomic composition and
diversity analysis at the genus level. Taxonomic and functional
trajectories of community composition were followed in a two-dimensional
NMDS ordination plan. Two NMDS using abundance-based (Bray-Curtis)
dissimilarity measures were conducted to map either taxonomic or
functional composition, the latter based on the seven leaf, stem and
life history traits (without seed mass classes). Trajectories along time
were reported through the Euclidean distance between the target
inventories and the reference inventories in 1989, \emph{i.e.} 2 years
after disturbance. Univariate trajectories of the leaf, stem and
life-history traits were also visualized with the community weighted
means (CWM) \citep{Diaz2007}. Species seed mass were given in 5 mass
classes. Seed mass trajectories were therefore reported as the
proportion of each class in the inventories (Supplementary materials).

The taxonomic diversity was reported through species richness and the
Hill number translation of the Simpson index \citep{Hill1973}. These
metrics allowed assessing the taxonomic richness as well as evenness,
through the comparison between richness and Simpson diversity. R.esults
will thus be discussed directly in terms of taxonomic richness and
evenness. Both indices belong to the set of HCDT or generalized entropy,
respectively corresponding to the 0 and 2 order of diversity (q),
recommended for diversity studies \citep{Marcon2015b}. The functional
diversity was reported using the functional richness and functional
evenness, \emph{i.e} the Rao index of quadratic entropy. The Rao index
combines species abundance distribution, and the average pairwise
functional dissimilarity between species computed by the gower distance.

@@

Functional redundancy was measured as the overlap among species in
community functional space \citep{Carmona2016}. First, the individuals
of the trait database were mapped in the plane of the two first axes
from a PCA analysis. Then, for each species, the traits probability
density (TPD) were computed through two-dimension kernel density
estimators. Second, for each community, the TDB weigthed by species
abundance were summed accross the functional space. Third, the
functional space was divided into a 100 x 100 grid and the number of
species with a positive TDP was counted in each cell. The average count
across cells minus 1 returnedthe Community Functional Redundancy, which
was the average number of species in the community that share the same
trait values.

@@

\caption{Plot trajectories in terms of taxonomic composition (\textbf{(a)} and \textbf{(c)}) and functional composition (\textbf{(b)} and \textbf{(d)}) in a two-dimensional NMDS plan. Lower panels (\textbf{(c)} and \textbf{(d)}) represent the Euclidean distance to initial condition along the 30 sampled years. Shaded areas are the credibility intervals.}\label{fig:NMDSplans}
\end{figure*}

@@

The second-degree polynomial regressions between (i) the percentage AGB
loss and (ii) taxonomic and functional diversity after 10, 20 and 30
years best predicted the hump-shaped curve of the disturbance impact
along the disturbance intensity gradient \ref{fig:IDHplot}. The
relationship between the disturbance impact and its intensity was more
markedly hump-shaped for the taxonomic richness than for the Simpson
diversity. For both functional richness and Rao diversity the
relationship was almost linear. The regression model better predicted
the functional richness and Rao diversity
(\(0.55<R^2_{Functional Richness}<0.72\), and
\(0.60<R^2_{Functional Rao}<0.81\)) than the taxonomic richness and
evenness (\(0.21<R^2_{Taxonomic Richness}<0.4\), and
\(-0.15<R^2_{Taxonomic Simpson}<0.43\) respectively).

@@

The taxonomic differences among local communities were already apparent
before disturbance, as revealed by the distinct starting points on the
NMDS axis 2. These differences were maintained throughout recovery
trajectories.\\
More than commonly thought, post-disturbance trajectories depended on
community initial composition, which partly determined the pool of
recruited species and constrained the trajectories towards the initial
composition. The community taxonomy proved highly resilient, as it
absorbed the disturbance and maintained commnity initial composition
characteristics \citep{Folke2006}. This high resilience suggested that
species not belonging to the pre-disturbance community were rarely
recruited because of the commonness of dispersal limitation among
tropical tree species \citep{Svenning2005}.

@@

Community taxonomic richness increased after disturbance until an
intensity threshold (20-25\% AGB loss) above which the taxonomic
richness decreased. The increase of taxonomic evenness after disturbance
also depended on the disturbance intensity, although the correlation was
weaker than for the taxonomic richness as already observed in the Guiana
Shield \citep{Baraloto2012a} and in Bornean tropical forests
\citep{Cannon1998}. In accordance with the IDH, then, community
taxonomic trajectories depended on the disturbance intensity. They
markedly changed above an intensity threshold, with a decrease of the
richness and a persistent increase of the evenness. Disturbance
intensity determined the balance in the community between surviving
trees from pre-disturbance communities and trees recruited afterward.
The pool of true pioneer species specifically recruited after
disturbance is, in the Guiana Shield, restricted to a few common genera
(e.g. \emph{Cecropia} spp., \emph{Vismia} spp.)\citep{Guitet2018}. Below
the intensity threshold the surviving community trees remained numerous
enough to maintain the pre-disturbance high taxonomic richness while the
pioneers recruited, infrequent or absent before disturbance, increased
both community taxonomic richness and evenness. Beyond the intensity
threshold, disturbance decreased the taxonomic richness of surviving
trees, and this decrease was not offset by the recruitment of pioneers.
The overall community taxonomic richness therefore decreased according
to the disturbance intensity \citep{Molino2001}. An intermediate
disturbance was then detected, defined by an intensity for which
post-disturbance trajectories markedly changed. Such disturbance
increased the availability of resources and created opportunities for
pioneers, without implying too important loss of shade-tolerant species
or preventing their maintenance in the community \citep{Bongers2009}.
For community taxonomic evenness, the disturbance impact was similar but
milder. Because the evenness is less sensitive to the loss of rare
species, below the intensity threshold the trajectories of taxonomic
evenness rather represented the increasing dominance of pioneers that
balancing the usual hyper-dominance of a few species. Beyond the
intensity threshold, however, pioneers became in turn highly dominant
and decreased the overall evenness \citep{Baraloto2012a}.
